Safe Transferable Regions

### 2016-10-13 paper/type-inference.tex
minor

\begin{theorem}
\label{thm:constraint-generation-sc}
Let $C = \consOf{q}$.
Any assignment $\soln$ (for the region and predicate variables in $q$)
satisfies $C$ iff $q[\soln]$ is well-typed.
\end{theorem}

@@

For any predicate variable $\varphi$ used in $C$, $C$ has exactly one constraint
of the form $\tywf{\Delta}{\varphi}$. We will refer to this $\Delta$ as
$\predDeltaMap(\varphi)$.

The formalism below is motivated by the following intuition.
Suppose some construct in a method's body requires an outlives-constraint
$\pi_1 \outlives \pi_2$ to hold. If both $\pi_1$ and $\pi_2$ are region parameters
of the method, then we can incorporate the requirement $\pi_1 \outlives \pi_2$ 
as part of the method's precondition ($\varphi$). Otherwise, the outlives-constraint
$\pi_1 \outlives \pi_2$ must be established by the context within the method
(\eg, by the \code{letregion} constructs in the method).
We now generalize this idea to the setting of our abstract constraint-satisfication
problem.

 We refer to an abduction constraint in $C$ as \emph{decomposable}
if its antecedent is of the form $\varphi \conj \phictxt$ where $\varphi$ is a predicate variable and
$\phictxt$ is a conjunction of zero or more outlives-constraints of the form
$\pi_1 \outlives \pi_2$ satisfying the following conditions:
(1) $\pi_2 \notin \predDeltaMap(\varphi)$.
(2) if $\pi_1 \in \predDeltaMap(\varphi)$, then for
every $\pi_f \in \predDeltaMap(\varphi)$, $\pi_f \outlives \pi_2$ is
a conjunct in $\phictxt$.

@@

Note that the above lemma shows how the conjunction in the antecedent of
any decomposable constraint can be eliminated.

\paragraph{Constraint Solver.}
The first step in our algorithm for solving a set of constraints $C$ uses an
iterative fixed point algorithm to compute a $\satC \supseteq C$ of
constraints by adding  a set of constraints that are logically implied by $C$.
% \paragraph{Step 1.}
We compute $\saturate{C}$ by iteratively applying the following rules until a fixed
point is reached:

@@

% Point~\ref{item:context} makes use of Lemmas~\ref{lemma:gc-is-decomposable} and~\ref{lemma:decomposition}.

@@

One key point of the first step is that it has the effect of unifying region variables with the
region constants that they should be bound to in the solution, as explained below.

Consider a set of constraints $D$. We say that a region variable $\rho$
\emph{occurs} in a \emph{context} $\ell$ (in $D$) if $D$ contains some
constraint $\isvalid{\ell}{r}$ where $\rho$ occurs in $r$.
We say that a region variable $\rho$ is \emph{bound} to a region constant
$\pi$ in a context $\ell$ if
$\{ \isvalid{\ell}{\rho \outlives \pi}, \isvalid{\ell}{\pi \outlives \rho} \} \subseteq D$.
We say that a region variable $\rho$ is \emph{bound} to a region constant $\pi$ (in $D$) if
$\rho$ is bound to $\pi$ in every context $\ell$ in which it occurs.

@@

The set $\satC$ also makes it easy to check if $C$ is satisfiable.
Let $\myground{C}$ denote the subset of all ground constraints
(\ie, constraints without any region variable or predicate variable)
in $\saturate{C}$. Let $\rhoC$ denote the subset of all wellformedness
constraints for region variables in $\satC$.
Define $\solveCon(C)$ as below.
% It is possible to show that $\soln$ satisfies $\satC4
% iff  $\myground{C}$ is valid in the theory of partial-order.

@@

\paragraph{Algorithmic Aspects.}
Note that checking the validity of a ground constraint can be realized using
a simple graph reachability algorithm.
%
Given a set $S$ of outlives constraints, define the directed
graph $G(S)=(V(S),E(S))$ as follows.
Every distinct region identifier $\rho$ in $S$ is represented by a vertex,
which we will also refer to as $\rho$.
Every outlives constraint $\rho_1 \outlives \rho_2$ is represented by
an edge from $\rho_1$ to $\rho_2$.
(Recall that an equality constraint $\rho_1 = \rho_2$ is shorthand
for the pair of constraints $\rho_1 \outlives \rho_2$ and
$\rho_2 \outlives \rho_1$: thus, an equality constraint is represented
by a pair of edges.)
%
It is easy to see that $\isvalid{\conj S}{\rho_1 \outlives \rho_2}$ iff
there exists a path from $\rho_1$ to $\rho_2$ in $G(S)$.
%
Thus, a simple graph reachability algorithm can be used to check the validity of
ground constraints.
% all the logical consequences of a set of outlives constraints.

This idea generalizes. Consider a set $C$ of constraints that contains no
constraint with an antecedent $\varphi \conj \phi$ where $\phi$ is a non-empty conjunction. 
Then, the algorithm for computing the closure $\satC$ does not need to use
rule~\ref{item:context} (for Abduction Decomposition). The closure in this
case can be computed using context-free reachability in graphs~\cite{Reps:Reachability}
(as usual for context-sensitive interprocedural analysis).

Algorithms for context-free reachability can be adapted to incorporate
the Abduction Decomposition step.
% Further details omitted.
Alternatively, the iterative process described above is a standard fixed point computation.
We can encode this computation using a set of Datalog rules,
and compute the solution using any Datalog engine.

>>>>

\begin{theorem}
\label{thm:constraint-generation-sc}
Let $C = \consOf{q}$.
An assignment $\soln$ (for the region and predicate variables in $q$)
satisfies $C$ iff $q[\soln]$ is well-typed.
\end{theorem}

@@

For any predicate variable $\varphi$ used in $C$, $C$ has exactly one constraint
of the form $\tywf{\Delta}{\varphi}$. We will refer to this $\Delta$ as
$\predDeltaMap(\varphi)$.
%
%% The formalism below is motivated by the following intuition.
%% Suppose some construct in a method's body requires an outlives-constraint
%% $\pi_1 \outlives \pi_2$ to hold. If both $\pi_1$ and $\pi_2$ are region parameters
%% of the method, then we can incorporate the requirement $\pi_1 \outlives \pi_2$ 
%% as part of the method's precondition ($\varphi$). Otherwise, the outlives-constraint
%% $\pi_1 \outlives \pi_2$ must be established by the context within the method
%% (\eg, by the \code{letregion} constructs in the method).
%% We now generalize this idea to the setting of our abstract constraint-satisfication
%% problem.
%
We refer to an abduction constraint in $C$ as \emph{decomposable}
if its antecedent is of the form $\varphi \conj \phictxt$ where $\varphi$ is a predicate variable and
$\phictxt$ is a conjunction of zero or more outlives-constraints of the form
$\pi_1 \outlives \pi_2$ satisfying the following conditions:
(1) $\pi_2 \notin \predDeltaMap(\varphi)$.
(2) if $\pi_1 \in \predDeltaMap(\varphi)$, then for
every $\pi_f \in \predDeltaMap(\varphi)$, $\pi_f \outlives \pi_2$ is
a conjunct in $\phictxt$.
%
(The above concepts parametrically depend on the set $C$, though we omit $C$
from the notation for simplicity.)

@@

The above lemma shows how we can reduce a non-trivial abduction constraint to
either a trivial abduction constraint or a non-abduction constraint.

\paragraph{Constraint Solver.}
The first step in our algorithm for solving a set of constraints $C$
computes a set $\satC \supseteq C$ of constraints by iteratively applying
the following rules until a fixed point is reached:

@@

% Point~\ref{item:context} makes use of Lemmas~\ref{lemma:gc-is-decomposable} and~\ref{lemma:decomposition}.
We can show that every new constraint added by the above rules is implied by existing
constraints:

@@

A key goal of this step is to identify the value every region variable must
have in any solution of the set of constraints, as explained below.
%
% unifying region variables with the
% region constants that they should be bound to in the solution, as explained below.
%
Consider a set of constraints $D$. We say that a region variable $\rho$
\emph{occurs} in a \emph{context} $\ell$ (in $D$) if $D$ contains some
constraint $\isvalid{\ell}{r}$ where $\rho$ occurs in $r$.
We say that a region variable $\rho$ is \emph{bound} to a region constant
$\pi$ in a context $\ell$ if
$\{ \isvalid{\ell}{\rho \outlives \pi}, \isvalid{\ell}{\pi \outlives \rho} \} \subseteq D$.
We say that a region variable $\rho$ is \emph{bound} to a region constant $\pi$ (in $D$) if
$\rho$ is bound to $\pi$ in every context $\ell$ in which it occurs.

@@

The set $\satC$ also makes it easy to check if $C$ is satisfiable.
Let $\myground{C}$ denote the subset of all ground constraints
(\ie, constraints without any region variable or predicate variable)
in $\saturate{C}$. Let $\rhoC$ denote the subset of all wellformedness
constraints for region variables in $\satC$.
Define $\solveCon{C}$ as below.
% It is possible to show that $\soln$ satisfies $\satC4
% iff  $\myground{C}$ is valid in the theory of partial-order.

@@

\paragraph{Algorithmic Aspects.}
Note that checking the validity of a ground constraint can be realized using
a simple graph reachability algorithm.
%
Given a set $S$ of outlives constraints, define the directed
graph $G(S)=(V(S),E(S))$ as follows.
Every distinct region identifier $\rho$ in $S$ is represented by a vertex,
which we will also refer to as $\rho$.
Every outlives constraint $\rho_1 \outlives \rho_2$ is represented by
an edge from $\rho_1$ to $\rho_2$.
%
%% (Recall that an equality constraint $\rho_1 = \rho_2$ is shorthand
%% for the pair of constraints $\rho_1 \outlives \rho_2$ and
%% $\rho_2 \outlives \rho_1$: thus, an equality constraint is represented
%% by a pair of edges.)
%
It is easy to see that $\isvalid{\conj S}{\rho_1 \outlives \rho_2}$ iff
there exists a path from $\rho_1$ to $\rho_2$ in $G(S)$.
%
Thus, a simple graph reachability algorithm can be used to check the validity of
ground constraints.
% all the logical consequences of a set of outlives constraints.

This idea generalizes.
% Consider a set $C$ of constraints that contains no non-trivial abduction constraint.
% Then, the algorithm for computing the closure $\satC$ does not need to use
% rule~\ref{item:context} (for Abduction Decomposition).
Extending the simple graph reachability algorithm to incorporate the Substitution rule
(in computing $\satC$) turns the problem into a context-free reachability problem in graphs~\cite{Reps:Reachability}
(as usual for context-sensitive interprocedural analysis).

Algorithms for context-free reachability can be adapted to incorporate the Abduction Decomposition step.
Alternatively, the iterative process described above is a standard fixed point computation
and can be encoded using a set of Datalog rules, allowing us to compute the closure
using any Datalog engine.
% Further details omitted.

### 2016-10-13 paper/typeinference-proofs.tex
minor

\begin{theorem}
\label{thm:constraint-generation-sc}
Let $C = \consOf{q}$.
Any assignment $\soln$ (for the region and predicate variables in $q$)
satisfies $C$ iff $q[\soln]$ is well-typed.
\end{theorem}
\begin{proof}
The correspondence between the static semantics rules and the constraint generation
rules induces a correspondence between derivation trees produced by the
two systems.
% We do not present a formal proof of the theorem.
We can establish the following lemma inductively:
given a derivation tree $\zeta_1$ for
$\exprok {\stdcontext}{e}{\tau}{C}$
and any substitution $\soln$,
we can construct a corresponding \emph{candidate} derivation
tree $\zeta_2$ of $\hastyp {\A[\soln],\env[\soln]}{e[\soln]}{\tau[\soln]}$.
We can show that $\zeta_2$ is a valid derivation tree iff $\soln$ satisfies $C$.
(We make use of analogous lemmas for type well-formedness rules as well.)
\end{proof}

@@

\paragraph{Soundness and Completeness of the Constraint Solver.}
\begin{lemma}
Let $C = \consOf{q}$.
Suppose a region variable $\rho$ is bound to two different region constants $\pi_1$
and $\pi_2$ in a context $\varphi \conj \phi$. % in $\satC$.
Let $\soln$ satisfy $\satC$.
For any context $\varphi \conj \phi$ that contains $\rho$, we must have
$\{ \pi_1, \pi_2 \} \subseteq \predDeltaMap(\varphi)$ and
$\isvalid{\predSubstFn(\varphi)}{\pi_1 = \pi_2}$.
\end{lemma}

@@

\begin{lemma}
Let $C = \consOf{q}$.
Let $\satC_1$ denote the subset of $\satC$ obtained by removing all constraints
that contain a region variable.
$\satC$ is satisfiable iff $\satC_1$ is satisfiable.
\end{lemma}

\begin{proof}
  The forward implication is trivial.
  The reverse implication follows from Lemma~\label{lemma:completely-bound}, as shown below.

\begin{theorem}
\label{thm:constraint-solver-soundness}
Let $C = \consOf{q}$.
If the constraint solver returns an assignment $\soln$,
then $\soln$ satisfies $\satC$.
\end{theorem}

\begin{proof}
Assume that the constraint solver returns an assignment $\soln$.

By construction, $\groundC$ (the set of all variable-free constraints in $\satC$)
is satisfiable. Hence, $\soln$ trivially satisfies all constraints in $\groundC$.

The definition of $\soln$ ensures that it satisfies the well-formedness constraints
for any region variable $\rho$.

Consider any constraint of the form $\isvalid{\varphi}{\pi_i \outlives \pi_j}$ in $\satC$,
where $\pi_i$ and $\pi_j$ are region constants. The definition of $\predSubstFn(\varphi)$
ensures that $\soln$ satisfies this constraint.
Furthermore, $\soln$ must also satisfy the well-formedness constraint
$\tywf{predDeltaMap(\varphi)}{\varphi}$.
(Otherwise, it follows from abduction decomposition that $\isvalid{}{\pi_i \outlives \pi_j}$ must
be in $\satC$, which means $\groundC$ must be invalid.)

@@

Consider any constraint of the form $\isvalid{\ell}{r}$ where $r$ contains a single occurrence
of a region variable, say $\rho$.
It follows from Lemma~\label{lemma:completely-bound} that $\rho$ is bound to some region
constant $\pi$.
Let $r'$ denote $r$ with $\rho$ replaced by $\pi$.
It follows from the transitivity rule that $\satC$ must contain the constraint $\isvalid{\ell}{r'}$
and that $\soln$ satisfies this constraint.
Further, it follows that $\soln$ satisfies $\isvalid{\ell}{\pi = \pi'}$ where $\pi' = \regionSubstFn(\rho)$.
Hence, $\soln$ satisfies this constraint as well.
\TODO{Flesh this out.}

@@

\begin{proof}
If $\satC$ is satisfiable, then $\groundC$ is trivially satisfiable.
If $\groundC$ is satisfiable, the solver will fail to return an assignment only if the set $S_\rho$ = 
$\{ \pi \in \regionDeltaMap(\rho) \; | \; \isvalid{\ell}{\rho = \pi} \in \satC \}$ is empty for some region variable $\rho$.
Consider any such $\rho$.
Let $\soln$ be a satisfying assignment for $\satC$.
Let $\pi' = \predSubstFn(\rho)$. Then, we must have $\pi' in \regionDeltaMap(\rho)$.
From Lemma~\ref{lemma:completely-bound}, $\rho$ must be bound to some $\pi$ in some context $\ell$.
Thus, $\isvalid{\ell}{\rho = \pi} \in \satC$.
Since $\soln$ satisfies this constraint, we have $\isvalid{\ell[\soln]}{\pi' = \pi}$.
If $\pi$ and $\pi'$ are the same region identifier, we have a contradiction, since
$\pi \in S_\rho$ and $S_\rho$ cannot be empty.
If $\pi$ and $\pi'$ are distinct region identifiers, then it follows that the precondition of case (a) must hold
in Lemma~\ref{lemma:decomposition}, and we must have $\{ \pi, \pi' \} \subseteq \predDeltaMap(\varphi)$.
We can show, by induction over the constraint-generation rules, that if $\regionDeltaMap(\rho)$ includes
some element of $\predDeltaMap(\varphi)$, then it includes all elements of $\predDeltaMap(\varphi)$.
It follows that $\pi \in \regionDeltaMap(\rho)$. Hence, the set $S_\rho$ is non-empty (as it contains $\pi$.
The result follows.
\end{proof}

\begin{theorem}
\label{thm:constraint-solver-sc}
Let $C = \consOf{q}$.
(Soundness) If $\solveCon{C} = \textsc{Some}(\soln)$, then $\soln$ satisfies $C$.
(Completeness) If $\solveCon{C} = \textsc{None}$, then $C$ is unsatisfiable.
\end{theorem}

>>>>

\begin{theorem}
\label{thm:constraint-generation-sc}
Let $C = \consOf{q}$.
An assignment $\soln$ (for the region and predicate variables in $q$)
satisfies $C$ iff $q[\soln]$ is well-typed.
\end{theorem}
\begin{proof}
The correspondence between the static semantics rules and the constraint generation
rules induces a correspondence between derivation trees produced by the
two systems.
% We do not present a formal proof of the theorem.
We can establish the following lemma inductively:
given a derivation tree $\zeta_1$ for
$\exprok {\stdcontext}{e}{\tau}{C}$
and any substitution $\soln$,
we can construct a corresponding \emph{candidate} derivation
tree $\zeta_2$ of $\hastyp {\A[\soln],\env[\soln]}{e[\soln]}{\tau[\soln]}$.
We can show that $\zeta_2$ is a valid derivation tree iff $\soln$ satisfies $C$.
(We make use of analogous lemmas for type well-formedness rules as well.)
\end{proof}

@@

\begin{lemma}
  \label{lemma:two-bindings}
Let $C = \consOf{q}$.
Suppose a region variable $\rho$ is bound to two different region constants $\pi_1$
and $\pi_2$ in a context $\varphi \conj \phi$. % in $\satC$.
Let $\soln$ satisfy $\satC$.
For any context $\varphi \conj \phi$ that contains $\rho$, we must have
$\{ \pi_1, \pi_2 \} \subseteq \predDeltaMap(\varphi)$ and
$\isvalid{\predSubstFn(\varphi)}{\pi_1 = \pi_2}$.
\end{lemma}

@@

% \begin{lemma}
% Let $C = \consOf{q}$.
% Let $\satC_1$ denote the subset of $\satC$ obtained by removing all constraints
% that contain a region variable.
% $\satC$ is satisfiable iff $\satC_1$ is satisfiable.
% \end{lemma}
% 
% \begin{proof}
%   The forward implication is trivial.
%   The reverse implication follows from Lemma~\label{lemma:completely-bound}, as shown below.
% \end{proof}

\begin{theorem}
\label{thm:constraint-solver-soundness}
Let $C = \consOf{q}$.
If $\solveCon{C} = \textsc{Some}(\soln)$,
then $\soln$ satisfies $\satC$.
\end{theorem}

\begin{proof}
Assume $\solveCon{C} = \textsc{Some}(\soln)$.

By definition (of $\solveCon{C}$), $\groundC$ (the set of all variable-free constraints in $\satC$)
is satisfiable. Hence, $\soln$ trivially satisfies all constraints in $\groundC$.
%
The definition also ensures that $\soln$ satisfies the well-formedness constraints
for any region variable $\rho$.

Consider any constraint of the form $\isvalid{\varphi}{\pi_i \outlives \pi_j}$ in $\satC$,
where $\pi_i$ and $\pi_j$ are distinct region constants. The definition of $\predSubstFn(\varphi)$
ensures that $\soln$ satisfies this constraint.
Furthermore, $\soln$ must also satisfy the well-formedness constraint
$\tywf{\predDeltaMap(\varphi)}{\varphi}$.
(Otherwise, the application of Abduction Decomposition to this constraint will add
$\isvalid{}{\pi_i \outlives \pi_j}$ to $\satC$. This constraint is invalid, which
contradicts the fact that $\groundC$ is valid.)

@@

Consider any constraint of the form $\isvalid{\ell}{r}$ where $r$ contains a single occurrence
of a region variable, say $\rho$.
It follows from Lemma~\ref{lemma:completely-bound} that $\rho$ is bound to some region
constant $\pi$.
Let $r'$ denote $r$ with $\rho$ replaced by $\pi$.
It follows from the transitivity rule that $\satC$ must contain the constraint $\isvalid{\ell}{r'}$.
Since this constraint has no variables on the consequent, we have already shown that $\soln$
must satisfy this constraint ($\isvalid{\ell}{r'}$).
Let $\pi' = \regionSubstFn(\rho)$.
We can show (as in the proof of Lemma~\ref{lemma:two-bindings}) that $\soln$ satisfies $\isvalid{\ell}{\pi = \pi'}$.
Hence, $\soln$ satisfies $\isvalid{\ell}{r}$ as well.

@@

\begin{proof}
If $\satC$ is satisfiable, then $\groundC$ is trivially satisfiable.
If $\groundC$ is satisfiable, the solver will fail to return an assignment only if the set $S_\rho$ = 
$\{ \pi \in \regionDeltaMap(\rho) \; | \; \isvalid{\ell}{\rho = \pi} \in \satC \}$ is empty for some region variable $\rho$.
Consider any such $\rho$.
Let $\soln$ be a satisfying assignment for $\satC$.
Let $\pi' = \predSubstFn(\rho)$. Then, we must have $\pi' \in \regionDeltaMap(\rho)$.
From Lemma~\ref{lemma:completely-bound}, $\rho$ must be bound to some $\pi$ in some context $\ell$.
Thus, $\isvalid{\ell}{\rho = \pi} \in \satC$.
Since $\soln$ satisfies this constraint, we have $\isvalid{\ell[\soln]}{\pi' = \pi}$.
If $\pi$ and $\pi'$ are the same region identifier, we have a contradiction, since
$\pi \in S_\rho$ and $S_\rho$ cannot be empty.
If $\pi$ and $\pi'$ are distinct region identifiers, then it follows that the precondition of case (a) must hold
in Lemma~\ref{lemma:decomposition}, and we must have $\{ \pi, \pi' \} \subseteq \predDeltaMap(\varphi)$.
We can show, by induction over the constraint-generation rules, that if $\regionDeltaMap(\rho)$ includes
some element of $\predDeltaMap(\varphi)$, then it includes all elements of $\predDeltaMap(\varphi)$.
It follows that $\pi \in \regionDeltaMap(\rho)$. Hence, the set $S_\rho$ is non-empty (as it contains $\pi$.
The result follows.
\end{proof}

\begin{theorem}
\label{thm:constraint-solver-sc}
Let $C = \consOf{q}$.\\
(Soundness) If $\solveCon{C} = \textsc{Some}(\soln)$, then $\soln$ satisfies $C$.\\
(Completeness) If $\solveCon{C} = \textsc{None}$, then $C$ is unsatisfiable.
\end{theorem}
